---
title: "Automated RNA-seq analysis of project `{r} params$project`"
author: "Vinicius M. Rios"
date: last-modified
format: 
  html:
    grid:
      body-width: 1000px
    toc: true
    code-fold: false
    code-tools: true
    theme: spacelab
execute:
  warning: false
  cache: refresh
  execute-dir: project
  echo: false
params:
  project: "PRJNA427332"
---


## 1. Introduction



------------------------------------------------------------------------

## 2. General statistics 

```{r delete_for_final_version}
#| echo: false
# Just to run in Vs-code, remove later
# params<-list(project= "PRJNA427332", zero=0)
# params<-list(project= "PRJNA376873", zero=0)

```

```{r load_libraries}
library(tidyverse)
library(GenomicFeatures)
library(txdbmaker)
library(tximport)
library(DESeq2)
library(org.Hs.eg.db)
library(RColorBrewer)
library(pheatmap)
library(patchwork)
library(ComplexUpset)
```


```{r get_files_path}
files_path <- list.files(paste0("results/quantification/salmon/", params$project),
  pattern = "quant.sf",
  recursive = TRUE,
  full.names = TRUE)

files_names <- dirname(files_path) %>%
  basename()

names(files_path) <- files_names

```

There are `{r} length(files_path)` files in the project `{r} params$project`.


```{r get_tx2gene}
gtf <- "references/homo_sapiens.gtf.gz"

txdb <- txdbmaker::makeTxDbFromGFF(gtf)

k <- biomaRt::keys(txdb, keytype = "TXNAME")

tx2gene <- biomaRt::select(txdb,
  keys = k,
  columns = "GENEID",
  keytype = "TXNAME")

colnames(tx2gene) <- c("TXNAME", "GENEID")

```

```{r tximport}
txi <- tximport(
  files = files_path,
  type = "salmon",
  tx2gene = tx2gene,
  countsFromAbundance = "lengthScaledTPM",
  ignoreTxVersion = TRUE
)
```

A total of `{r} prettyNum(nrow(tx2gene), big.mark = ",")` transcripts were mapped to `{r} prettyNum(nrow(txi$counts),big.mark = ",")` genes.

```{r get_metadata}
metadata <- readr::read_tsv(paste0("data/", params$project, ".metadata.clean"))
```

```{r DESeqDataSet}
dds <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ treatment  
)
```

```{r library_barplot}
#| fig-width: 16
#| fig-height: 6
library_sizes <- colSums(assay(dds))

library_df <- data.frame(Library_Size = library_sizes) %>%
  rownames_to_column(.) %>%
  left_join(x=., y = metadata, by = c("rowname" = "run_accession")) %>%
  mutate(sample_treatment = paste(sample,treatment))

ggplot(data = library_df,
       mapping = aes(x = rowname,
                     y = Library_Size,
                     fill = treatment)) +
  geom_col() +
  geom_hline(yintercept=mean(library_df$Library_Size), linetype = "dashed") +
  labs(x = "Run Accession",
       y = "Library Size",
       title = "Library Size by Run Accession",
       subtitle = "With mean library size line")+
  scale_fill_brewer(palette = "Set2")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```

```{r histogram}
#| fig-width: 16
smallest_group_size <- metadata %>%
    dplyr::count(treatment)%>%
    summarise(min_group_size = min(n)) %>%
    dplyr::pull(min_group_size)

genes_expressed <- (assay(dds) >= 10) %>%
  rowSums(.) %>%
  as.data.frame(.)%>%
  rownames_to_column(.)
  
colnames(genes_expressed) <- c("Genes","Count")

genes_expressed <- genes_expressed %>%
  mutate(keep_gene = ifelse(Count >= smallest_group_size, "Filtered", "Discarded"))

ggplot(data = genes_expressed, aes(x = Count, fill = keep_gene))+
  geom_histogram(bins = length(files_path)+1)+
  scale_fill_brewer(palette = "Paired")+
  labs(x = "Sample Count",
       y = "Number of genes",
       title = "Count of genes expressed in number of samples",
       fill = "Filtering Status")+
  scale_x_continuous(breaks = seq(0, length(files_path), by = 1))
```

```{r pre_filtering}
keep <- rowSums(counts(dds) >= 10) >= smallest_group_size
dds <- dds[keep, ]
levels_treatment <- levels(dds$treatment)
```

After filtering for counts >= 10 in at least `{r} smallest_group_size` samples (smallest group size) `{r} prettyNum(sum(keep), big.mark =",")` genes remained (`{r} 100*signif(sum(keep)/nrow(txi$counts), digits=3)`%)

```{r differential_expression_analysis}
dds <- DESeq(dds)

vsd <- vst(dds, blind = FALSE)
```

```{r pheatmap}
#| fig-width: 12
#| fig-height: 8
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$treatment, vsd$sample, sep="-")
colnames(sampleDistMatrix) <- NULL

colors <- colorRampPalette( rev(brewer.pal(9, "Oranges")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         main = "Sample-to-Sample Distances",
         col = colors)
```


```{r}
plotDispEsts(dds)

plotPCA(vsd, intgroup = "treatment") +
  theme_minimal() +
  ggtitle("PCA Plot of Samples")
```

```{r create_empty_results_list}
results <- list()
results_shrink <- list()
results_shrink_df <- list()
degs <- list()
```

```{r populate_results}

FDR_threshold <- 0.05
lfc_threshold <- 1

reference_list <- levels(dds$treatment)
for (i in seq_len(length(reference_list) - 1)) {
  for(comparison in DESeq2::resultsNames(dds)[-1]){
    results[[comparison]] <- DESeq2::results(dds, name = comparison)

    results_shrink[[comparison]] <- DESeq2::lfcShrink(
      dds,
      coef = comparison,
      type = "apeglm"
    )

    results_shrink_df[[comparison]] <- as.data.frame(results_shrink[[comparison]]) %>%
      tibble::rownames_to_column("gene_id") %>%
      arrange(padj)
    
    results_shrink_df[[comparison]]$symbol <- mapIds(org.Hs.eg.db,
      keys = results_shrink_df[[comparison]]$gene_id,
      column = "SYMBOL",
      keytype = "ENSEMBL",
      multiVals = "first"
    )

    results_shrink_df[[comparison]]$gene_name <- mapIds(org.Hs.eg.db,
      keys = results_shrink_df[[comparison]]$gene_id,
      column = "GENENAME",
      keytype = "ENSEMBL",
      multiVals = "first"
    )

    degs[[comparison]] <- results_shrink_df[[comparison]] %>%
      filter(padj < FDR_threshold & abs(log2FoldChange) > lfc_threshold)

  }
  keep <- dds$treatment != reference_list[i]
  dds <- dds[,keep]
  dds$treatment <- droplevels(dds$treatment)
  if (length(levels(dds$treatment))>1){
    dds <- DESeq(dds)
  }
}
```

```{r def_extract_deseq_summary}
extract_deseq_summary <-function(results, alpha =0.05){
  capture <- capture.output(summary(results, alpha = alpha))

  total_genes <- parse_number(capture[2])

  up_reg <- str_extract_all(capture[4],"\\d+(\\.\\d+)?")%>%
    unlist %>%
    as.numeric %>%
    .[2]

  down_reg <- str_extract_all(capture[5],"\\d+(\\.\\d+)?")%>%
    unlist %>%
    as.numeric %>%
    .[2]

  outliers <- str_extract_all(capture[6],"\\d+(\\.\\d+)?")%>%
    unlist %>%
    as.numeric %>%
    .[2]

  low_counts <- str_extract_all(capture[7],"\\d+(\\.\\d+)?")%>%
    unlist %>%
    as.numeric %>%
    .[2]

  summary_df <- list(
    "total_genes" = total_genes,
    "up_regulated" = up_reg,
    "up_regulated_percent" = up_reg/total_genes,
    "down_regulated" = down_reg,
    "down_regulated_percent" = down_reg/total_genes,
    "outliers" = outliers,
    "outliers_percent" = outliers/total_genes,
    "low_counts" = low_counts,
    "low_counts_percent" = low_counts/total_genes
  )
  return(summary_df)
}
```

```{r}

```

## 3. DEG analysis of pairwise comparisons
::: {.panel-tabset group="treatment"}

```{r print_summary}
#| echo: false
#| output: asis


for (name in names(results)) {
  formated_name <- name %>%
    str_remove("^treatment_") %>%
    str_replace_all("_", " ")
  
  cat("###", formated_name, "\n\n")

  deseq_summary <- extract_deseq_summary(results[[name]])

  deseq_summary_shrink <- extract_deseq_summary(results_shrink[[name]])

  cat(paste0("**Total genes:** ", deseq_summary_shrink$total_genes,"\n\n"))

  cat(paste0("**Up regulated:** ",
    deseq_summary_shrink$up_regulated,
    " (",
    formatC(signif(deseq_summary_shrink$up_regulated_percent, digits=3), digits=3, format="fg", flag="#"),
    "%)",
    "\n\n")
  )

  cat(paste0("**Down regulated:** ",
    deseq_summary_shrink$down_regulated,
    " (",
    formatC(signif(deseq_summary_shrink$down_regulated_percent, digits=3), digits=3, format="fg", flag="#"),
    "%)",
    "\n\n")
  )

  cat(paste0("**Outliers:** ",
    deseq_summary_shrink$outliers,
    " (",
    formatC(signif(deseq_summary_shrink$outliers_percent, digits=3), digits=3, format="fg", flag="#"),
    "%)",
    "\n\n")
  )

  cat(paste0("**Low counts:** ",
    deseq_summary_shrink$low_counts,
    " (",
    formatC(signif(deseq_summary_shrink$low_counts_percent, digits=3), digits=3, format="fg", flag="#"),
    "%)",
    "\n\n")
  )

  cat("\n\n")
}
```

:::

```{r upset_plot}
#| fig-width: 16
ven_data <- list()
for (name in names(results_shrink_df)){
  gene_vector  <- results_shrink_df[[name]] %>%
    drop_na(padj) %>%
    dplyr::pull(gene_id)

  ven_data[[name]] <- gene_vector 
}
all_genes <- unique(unlist(ven_data))

ven_matrix <- sapply(ven_data, function(x) all_genes %in% x)
ven_df <- as.data.frame(ven_matrix)
rownames(ven_df) <- all_genes
names(ven_df) <- names(ven_df)%>%
        str_remove("^treatment_") %>%
        str_replace_all("_", " ")


upset_plot <-  upset(data = ven_df,
        intersect = names(ven_df),
        intersections='all',
        set_sizes = FALSE,
        width_ratio=0.2,
        height_ratio=0.8,
        sort_sets = FALSE,
        name = "",
        base_annotations=list(
          'Intersection size'=intersection_size(
              counts=TRUE,
              mapping=aes(fill='bars_color'),
              text_colors=c(
                on_background='black', on_bar='black'
            )
          )+
          scale_fill_brewer(palette = "Paired", guide='none')+
          ylab('Gene count')+
          theme(axis.title.y = element_text(size = 18, ,vjust=0.5),
          axis.title.y.right = element_text(size = 18, vjust=0.5, margin = margin(l = 10)))+
          scale_y_continuous(position = 'right')
        ),
        wrap=TRUE,
        matrix=(
            intersection_matrix(
                geom=geom_point(
                    shape='square',
                    size=8,
                ),
                segment=geom_segment(
                  color = "#1f78b4",
                  linewidth = 1.2
                ),
                outline_color = list(active = "#1f78b4", inactive = NA)             
            )
        )+
        scale_color_manual(
            values=c('TRUE'='#1f78b4', 'FALSE'='grey'),
            guide = 'none'       
            ),
        themes = upset_default_themes(text = element_text(size = 15))
  )+
  ggtitle('Overlap between all genes')

ven_data_2 <- list()
for (name in names(results_shrink_df)){
  gene_vector  <- results_shrink_df[[name]] %>%
    drop_na(padj) %>%
    filter(padj <=0.05)%>%
    dplyr::pull(gene_id)

  ven_data_2[[name]] <- gene_vector 
}
all_genes_2 <- unique(unlist(ven_data_2))

ven_matrix_2 <- sapply(ven_data_2, function(x) all_genes_2 %in% x)
ven_df_2 <- as.data.frame(ven_matrix_2)
rownames(ven_df_2) <- all_genes_2
names(ven_df_2) <- names(ven_df_2)%>%
        str_remove("^treatment_") %>%
        str_replace_all("_", " ")


upset_plot_2 <-  upset(data = ven_df_2,
        intersect = names(ven_df_2),
        intersections='all',
        set_sizes = FALSE,
        width_ratio=0.2,
        height_ratio=0.8,
        sort_sets = FALSE,
        name = "",
        base_annotations=list(
          'Intersection size'=intersection_size(
              counts=TRUE,
              mapping=aes(fill='bars_color'),
              text_colors=c(
                on_background='black', on_bar='black'
            )
          )+
          scale_fill_brewer(palette = "Paired", guide='none')+
          ylab('DEGs count')+
          theme(axis.title.y = element_text(size = 18, ,vjust=0.5))+
          scale_y_continuous(position = 'right')  
        ),
        wrap=TRUE,
        matrix=(
            intersection_matrix(
                geom=geom_point(
                    shape='square',
                    size=8
                ),
                segment=geom_segment(
                  color = "#1f78b4",
                  linewidth = 1.2
                ),
                outline_color = list(active = "#1f78b4", inactive = NA)           
            )
        )+
            scale_color_manual(
            values=c('TRUE'='#1f78b4', 'FALSE'='grey'),
            guide = 'none'       
            ),
        themes = upset_default_themes(text = element_text(size = 15)),
        labeller = ggplot2::as_labeller(function(x) '')
  )+
  ggtitle('Overlap between DEGs')

upset_plot + upset_plot_2
```

## Plots

::: {.panel-tabset group="treatment"}

```{r ma_plot}
#| echo: false
#| output: asis
#| fig-width: 16
#| fig-height: 10

for (name in names(results)) {
  formated_name <- name %>%
    str_remove("^treatment_") %>%
    str_replace_all("_", " ")
  
  cat("###", formated_name, "\n\n")
  
  
  cat("\n\n")
  cat("#### MA Plot\n\n")

  volcano_data <- results_shrink_df[[name]] %>%
    mutate(deg = case_when(
      padj <= FDR_threshold & log2FoldChange > lfc_threshold ~ "Upregulated",
      padj <= FDR_threshold & log2FoldChange < -lfc_threshold ~ "Downregulated",
      TRUE ~ "Not-significant"
      )
    ) %>%
    mutate(significant = case_when(
      padj <= FDR_threshold ~"Significant",
      padj > FDR_threshold ~"Non-significant",
      is.na(padj) & !is.na(pvalue) ~ "Low normalized count",
      TRUE ~ "Outlier"
      )
    ) %>%
    mutate(y_display = case_when(
      log2FoldChange > 4 ~ 4,
      log2FoldChange < -4 ~ -4,
      TRUE ~ log2FoldChange
    ),
    shape_type = case_when(
      
      log2FoldChange > 4 ~ "upper",
      log2FoldChange < -4 ~ "lower",
      TRUE ~ "normal"
    )
  )

  ma_plot <- ggplot(volcano_data, aes(x = baseMean, y = log2FoldChange, color = significant, alpha = significant)) +
    geom_point(data = subset(volcano_data, shape_type == "normal"), 
              size = 2) +
    geom_point(data = subset(volcano_data, shape_type == "upper"),
              shape = 17, size = 2.5) +  # upward triangle
    geom_point(data = subset(volcano_data, shape_type == "lower"),
             shape = 6, size = 2.5) + 
    scale_x_log10() +
    theme_minimal() +
    labs(x = expression(log[10]~"(Mean of Normalized Counts)"), y = expression("Shrunken"~log[2]~"Fold Change"))+
    ylim(c(-5, 5))+
    theme(legend.title = element_blank(),
          legend.position = "top")+
    scale_color_manual(values = c("Significant" = "red", "Non-significant" = "grey", "Low normalized count" = "#895129", "Outlier" = "#8B008B"))+
    scale_alpha_manual(values = c("Significant" = 0.8, "Non-significant" = 0.3, "Low normalized count" = 0.5, "Outlier" = 0.5))+
    guides(color = guide_legend(override.aes = list(shape = 16)))

 

  volcano_plot <- ggplot(volcano_data, aes(x = log2FoldChange, y = -log10(padj), color = deg, alpha = deg)) +
    geom_point(size = 2) +
    scale_color_manual(values = c("Upregulated" = "red", 
                                  "Downregulated" = "#6960dbff", 
                                  "Not significant" = "gray")) +
    scale_alpha_manual(values = c("Upregulated" = 0.8, 
                                  "Downregulated" = 0.8, 
                                  "Not significant" = 0.5))+
    geom_vline(xintercept = c(-lfc_threshold, lfc_threshold), linetype = "dashed") +
    geom_hline(yintercept = -log10(FDR_threshold), linetype = "dashed") +
    labs(x = expression(log[2]~"Fold Change"),
        y = expression(-log[10]~"Adjusted P-value")) +
    theme_minimal()+
    theme(legend.title = element_blank(),
          legend.position = "top")

  

  print(ma_plot)

  cat("\n\n")
  cat("#### Volcano Plot\n\n")

  print(volcano_plot)

  cat("\n\n")
}

```

:::































